Physcraper: a python package for continual update of evolutionary estimates using the Open Tree of Life

### 2021-01-07 2.1-intro-ideas.Rmd
discussion
intro again
intro and discussion

>>>>

Connecting existing phylogenies with data such as
geographical location, fossil ranges, and genetic and phenotypic information increasingly
available in public databases constitute would be a great progress for evolutionary studies.

However, if there are multiple alignments and trees, TreeBASE does not always indicate how they are connected. Hence, Phylesystem's metadata only provides a loose means of linking the tree with the exact alignment that generated it and a human curator must make the link.

### 2021-01-07 sources/bmc_template_submission/bmc_article.tex
conclusion
intro and discussion

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Phylogenies capture the shared history of organisms and provide key evolutionary
context for our biological observations \cite{dobzhansky1973nothing}.
Public biological databases constitute an amazing resource for evolutionary
studies. Connecting existing phylogenies to molecular data that has never been
incorporated into any phylogenetic estimate, geographical location, fossils, and
other data in a reproducible and continuous manner is possible by establishing a
data interoperability framework for biological databases. Here, we introduce
Physcraper, a tool that automates biological database connections
with the main goal of building upon homology
hypotheses that taxon specialists have assessed and deemed appropriate for a
specific phylogenetic scope to update a starting tree and single locus alignments
with public DNA data, with the added capability to connect the lineages represented
in the phylogeny to biological data in public databases.

One of the main challenges for automatic
integration of biological data into phylogenies are varying taxonomic idiosyncrasies
across databases. To address this challenge, the Open Tree of Life project (OpenTree)
unifies taxonomic
data from various databases \cite{rees2017automated}, including the USA National
Center for Biodiversity Information (NCBI) molecular database GenBank
\cite{benson2000genbank}, among others. The OpenTree taxonomy represents a key resource
that remains to be leveraged for connecting data from any biological database that
has been integrated to it.

The Open Tree of Life project (OpenTree)
also constructs a comprehensive tree of life by synthesizing published phylogenies
and taxonomy. OpenTree's "synthetic" tree comprises 2.3 million tips, of which
around 90,000 (87,740) are supported by phylogenies - the remaining 1.4 million taxa are
placed in the tree based on taxonomy. Meaning that there is still a lot of work
to do to complete a full phylogenetic tree of all life.

@@

With more than XXXX million single locus
sequences in GenBank today, there is a considerable amount of phylogenetically
informative data in GenBank that has not been analysed and/or made publicly available
with the potential to give phylogenetic light to the tree of life.

Phylogenies constructed from single locus super mayrix datasets are still widely
generated and relevant to the field. Notable examples include the

Notable examples of phylogenies constructed using multi-locus, database mined, super matrix approach:
Magallon, Soltis, examples with fish, bats,
\cite{magallon2015metacalibrated} mining by hand - flowering plant families - has been cited 561 times
\cite{smith2018constructing} 38k clades of plants with seeds - cited 176 times
\cite{jetz2012global} 6k bird species - cited
\cite{rabosky2018inverse} mining with phlawd - cited 202 times

These studies are widely cited and immensely relevant for the biological community.

Some of these multilocus phylogenetic studies have implemented pipelines such as PHLAWD,
other studies have curated their matrices by hand, which is largely non-reproducible.

Make emphasis on these really great studies (and pipelines) that do not rely on
genomic data, and that there is a lot of single locus data that can be analysed.

Assembling a DNA alignment from such a massive database can be done "by hand",
but it is a largely time consuming and mostly non-reproducible approach.
Computational pipelines that mine DNA databases fast, efficiently, and reproducibly,
have been applied to infer phylogenetic relationships in a variety of organisms
(e.g., \cite{smith2009mega, antonelli2017toward, izquierdo2014pumper}).
However, fine-grained curated markers and alignments can improve phylogenetic
reconstructions, even in phylogenomic analyses \cite{fragoso2017pilot}.

There are almost 8,200 publicly available, peer-reviewed alignments, covering
around 100,000 distinct taxa in the TreeBASE database \cite{piel2009treebase}, which
can be used as seeds to mine molecular databases, and as "jump-start" alignments
for phylogenetic reconstructions \cite{morrison2006multiple} to continually enrich,
update and compare existing phylogenetic knowledge.

----------

Physcraper is a Python pipeline using OpenTree's taxonomy and programmatic access
protocols (API's) to implement a database interoperability framework that automatically
links phylogenies that have been standardized to OpenTree taxonomy, to alignments
from TreeBASE, data from GenBank, and phylogenies from OpenTree's Phylesystem.
Physcraper aims to demonstrate the benefits of reproducible workflows and open
science in phylogenetics, and encourage better data sharing practices in the community.

@@

Databases preserving and democratizing access to biological data
have become essential resources for science.
New molecular data keep accumulating and tools facilitating its integration into
existent evolutionary knowledge are needed.

Phylogenetic pipelines designed to make evolutionary sense of the vast amount of
public molecular data (e.g., Phylota \cite{sanderson2008phylota},
PHLAWD \cite{smith2009mega},
SUPERSMART \cite{antonelli2017toward}) focus on generating full phylogenies \textit{de novo},
i.e., inferring phylogenetic relationships from a newly generated homology hypothesis,
as opposed to e.g., supertrees, that are generated by summarizing previous
phylogenetic estimates.
While Physcraper does not generate phylogenies \textit{de novo} in a traditional sense,
it successfully generates new phylogenetic knowledge, revealing the importance of
open science in facilitating phylogenetic placement of public molecular data and
accelerating enrichment and updating of phylogenetic relationships in any region
of the tree of life.
The PUMPER pipeline \cite{izquierdo2014pumper} also uses the concept of updating
pre-existing alignments to incorporate public molecular data into phylogenies.
Unfortunately, installation was unsuccessful following instructions from the author,
and a comparison analysis between Physcraper and PUMPER is unfeasible at present.
%% **MTH: add some more details about what aspect of the PUMPER install failed**

@@

Data repositories hold more information than meets the eye.
Beyond the main data, they are rich sources of metadata that can be leveraged
for the advantage of all areas of biology as well as the advancement of scientific
policy and applications.
Initial ideas about the data are constantly changed by results from new analyses.
Physcraper provides a framework for reproducible phylogenetics that has the
potential to consistently provide context for these ideas, highlighting the
importance of data sharing and open science in the field, biology and science.

\section*{Conclusions}
Text for this section\ldots

>>>>

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Phylogenies capture the shared history of organisms and provide key evolutionary
context for our biological observations \cite{dobzhansky1973nothing}. Data such as
geographical location, fossil ranges, and genetic and phenotypic information increasingly
available in public databases constitute an amazing
resource for biological discovery \cite{baxevanis2015importance}.
Connecting existing phylogenies with public biological data in a reproducible and
continuous manner would represent a key tool for evolutionary studies.

Here, we introduce Physcraper, a tool that automates biological database connections
with the main goal of building upon homology
hypotheses that taxon specialists have assessed and deemed appropriate for a
specific phylogenetic scope to update a starting tree and single locus alignments
with public DNA data, with the added capability to connect the lineages represented
in the phylogeny to biological data from public databases.

One of the main challenges for automatic
integration of biological data into phylogenies are varying taxonomic idiosyncrasies
across databases. To address this challenge, the Open Tree of Life project (OpenTree)
created a unified taxonomy for name standardization, by integrating taxonomic
data from several databases \cite{rees2017automated}, including the USA National
Center for Biodiversity Information (NCBI) molecular database GenBank
\cite{benson2000genbank}, among others.
Existing OpenTree taxonomy programmatic tools
are a key resource that can be used to establish a framework for connecting data from any
biological database that has been integrated to OpenTree's unified taxonomy.

OpenTree's unified taxonomy is used to construct OpenTree's comprehensive tree of
life that synthesizes published phylogenies and taxonomy. It comprises 2.3 million tips, of which
around 90,000 (87,740) are supported by phylogenies - the remaining 1.4 million taxa are
placed in the tree based on taxonomy.
This large gap in public phylogenetic knowledge is due to several factors:
- lack of data sharing, most analyses essentially lost forever;
- lack of data, we still lack genetic data for a large portion of organisms; and
- lack of analyticla power of existing data.

@@

With more than 221 million single locus
sequences in GenBank today \cite{genbankstats}, there is a considerable amount of phylogenetically
informative data in GenBank that has not been analysed and/or made publicly available
with the potential to significantly fill phylogenetic gaps in the tree of life.

Assembling a DNA alignment from such a massive database can be done "by hand",
but it is a largely time consuming and mostly non-reproducible approach.
Computational pipelines that mine DNA databases fast, efficiently, and reproducibly
have been applied to infer phylogenetic relationships in a variety of organisms
(e.g., \cite{smith2009mega, antonelli2017toward, izquierdo2014pumper}).
While genomics has, and will continue to, revolutionize phylogenetic inference,
the diversity of alternative genomic sequencing approaches implemented produce
largely non-overlapping homology hypotheses across taxa \cite{jones2016targeted},
creating challenges for phylogenetic reconstruction.
Phylogenomics addresses this problem by focusing on targeted capture of informative
regions \cite{andermann2020guide}.
However, fine-grained curated markers and alignments can significantly improve phylogenetic
reconstructions, even in phylogenomic analyses \cite{fragoso2017pilot}.

There are almost 8,200 publicly available, peer-reviewed curated alignments, covering
around 100,000 distinct taxa in the TreeBASE database \cite{piel2009treebase}, which
can be leveraged as seeds to mine molecular databases, and as "jump-start" alignments
for phylogenetic reconstructions \cite{morrison2006multiple} to continually enrich,
update and compare existing phylogenetic knowledge.

Such an approach was proposed in PUMPER but the community has not picked it up,
probably due to installation difficulties.
Physcraper extends this approach by allowing for interoperability.

Physcraper is implemented as a Python pipeline using OpenTree's taxonomy and programmatic access
protocols (API's) to implement a database interoperability framework that automatically
links phylogenies that have been standardized to OpenTree taxonomy, to alignments
from TreeBASE, data from GenBank, and phylogenies from OpenTree's Phylesystem.
Physcraper aims to demonstrate the benefits of reproducible workflows and open
science in phylogenetics, and encourage better data sharing practices in the community.

@@

Databases preserving and democratizing access to biological data
have become essential resources for science.
New molecular data keep accumulating and tools facilitating its integration into
existent evolutionary knowledge contribute to the acceleration of scientific discovery.

Physcraper is the first tool for interoperability in phylogenetic analyses (???)
As a tool for automatization of phylogenetic reconstruction from molecular databases,
we believe it presents several advantages over existing phylogenetic pipelines designed to
make evolutionary sense of the vast amount of
public molecular data. Phylota \cite{sanderson2008phylota},
and PHLAWD \cite{smith2009mega} are the first ones and have been cited and used
the most for actual phylogenetic reconstruction and mining of databases. These are
strong tools that will probably still have a great potential in their recent updated versions
(phylotar \cite{bennett2018phylotar} and PyPHLAWD \cite{smith2019pyphlawd}).
They focus on generating full phylogenies \textit{de novo},
i.e., inferring phylogenetic relationships from a newly generated homology hypothesis,
as opposed to e.g., supertrees, that are generated by summarizing previous
phylogenetic estimates, which requires more manual adjustment later and difficults
comparison with previous phylogenetic information.

SUPERSMART \cite{antonelli2017toward}) and PUMPER \cite{izquierdo2014pumper}

The PUMPER pipeline \cite{izquierdo2014pumper} also uses the concept of updating
pre-existing alignments to incorporate public molecular data into phylogenies.
Unfortunately, installation was unsuccessful following instructions from the author,
and a comparison analysis between Physcraper and PUMPER is unfeasible at present.
%% **MTH: add some more details about what aspect of the PUMPER install failed**

@@

\section*{Conclusions}
Data repositories hold more information than meets the eye.
Beyond the main data, they are rich sources of metadata that can be leveraged
for the advantage of all areas of biology as well as the advancement of scientific
policy and applications.
Initial ideas about the data are constantly changed by results from new analyses.
Physcraper provides a framework for reproducible phylogenetics that has the
potential to consistently provide context for these ideas, highlighting the
importance of data sharing and open science in the field, biology and science.

### 2021-01-07 temp.Rmd
intro and discussion

While genomics has, and will continue to, revolutionize phylogenetic inference,
the diversity of alternative genomic sequencing approaches implemented produce
largely non-overlapping homology hypotheses across taxa \cite{jones2016targeted},
creating challenges for phylogenetic reconstruction.
Phylogenomics addresses this problem by focusing on targeted capture of informative
regions \cite{andermann2020guide}. Yet, it is expensive. Few labs around the world can afford
to implement it for all their study groups, or at all. Creating a wealth gap for
science, while free information is still available to provide new knowledge.
Genomics does not contribute yet to democratization of phylogenetic research \cite{li2017enabling}.

@@

>>>>

While Physcraper does not generate phylogenies \textit{de novo} in a traditional sense,
it successfully generates new phylogenetic knowledge, revealing the importance of
open science in facilitating phylogenetic placement of public molecular data and
accelerating enrichment and updating of phylogenetic relationships in any region
of the tree of life.

Genomics is expensive. Few labs around the world can afford
to implement it for all their study groups, or at all. Creating a wealth gap for
science, while free information is still available to provide new knowledge.
Genomics does not contribute yet to democratization of phylogenetic research \cite{li2017enabling}.

@@

%% While taxonomic trees are useful, trees inferred from evolutionary estimates are
preferred.

Notable examples of phylogenies constructed using multi-locus, database mined, super matrix approach:
Magallon, Soltis, examples with fish, bats,
\cite{magallon2015metacalibrated} mining by hand - flowering plant families - has been cited 561 times
\cite{smith2018constructing} 38k clades of plants with seeds - cited 176 times
\cite{jetz2012global} 6k bird species - cited
\cite{rabosky2018inverse} mining with phlawd - cited 202 times

These studies are widely cited and immensely relevant for the biological community.

Some of these multilocus phylogenetic studies have implemented pipelines such as PHLAWD,
other studies have curated their matrices by hand, which is largely non-reproducible.

Make emphasis on these really great studies (and pipelines) that do not rely on
genomic data, and that there is a lot of single locus data that can be analysed.
